---
pdf_document: default
title: "Decathlon PCA Exercise"
author: "Julianne Quinn"
abstract: 
output: pdf_document
---
***

```{r "setup", include=FALSE}
#This block of code will not appear in your knitted document
#Setup paths for data and Rcode
require("knitr")
sourcedir <-"~/Documents/4021/R Code"
opts_knit$set(root.dir = sourcedir)
opts_chunk$set(warning=FALSE)
```

Source files & load and clean data
```{r}
source("SPM_Panel.R")
source("PCAplots.R")

setwd(sourcedir)

library(ade4)

# load olympic dataset with decathlon data
data(olympic)

# create a data frame with the decathlon data
decathlon.data <- olympic$tab
decathlon.data$score <- olympic$score

# change French event names to English
names(decathlon.data)[names(decathlon.data) == "long"] <- "long_jump"
names(decathlon.data)[names(decathlon.data) == "poid"] <- "shot_put"
names(decathlon.data)[names(decathlon.data) == "haut"] <- "high_jump"
names(decathlon.data)[names(decathlon.data) == "110"] <- "110_hurdles"
names(decathlon.data)[names(decathlon.data) == "disq"] <- "discus"
names(decathlon.data)[names(decathlon.data) == "perc"] <- "pole_vault"
names(decathlon.data)[names(decathlon.data) == "jave"] <- "javelin"
  
```

1. Describe this dataset. What are the variables and what are the different observations? What relationships would you expect from this dataset?



2. Now look at your data using scatter plot matrices.
```{r}
pca <- princomp(decathalon.data[,c()])


```
3. Which features are strongly correlated? Which are most predictive of score?



4. Now create principal components with both the covariance and correlation matrices.
```{r}

```

5. Create a biplot of your data for both the covariance and correlation matrices
```{r}

```

6. What do you notice about the biplots with the 2 methods?



From here on out, use the correlation matrix.

7. Describe the relationships between the different variables. What do they imply about which decathlon events require similar or different strengths? (Note: In some events you want high scores, and in some events you want low scores.)



8. Create a screeplot.
```{r}

```

9. How many components do the graphs suggest is sufficient to explain most of the variation in the data? (Hint: consider the first method for choosing the number of PCs described in the slides.)



10. Create a cumulative variance plot.
```{r}

```

11. How many principal components do you need to explain at least 80% of the variance?



12. Plot the loadings in the first 2 PCS.
```{r}

```

13. Which 5 variables explain most of the variability in the first PC and how are they related to each other?



14. Which 5 variables explain most of the variability in the second PC and how are they related to each other?



15. Based on all of the analyses above, which events do you think are most important to an athlete's decathlon score?


